Aligning marine species range data to better serve science and conservation

### 2016-08-27 clip_depth/coral_maps.Rmd
pushing aquamaps border issues and manuscript updates

>>>>

``` {r plot_coral_composite_map}

library(tmap)
data(World)

coral_comp <- coral_cells_iucn %>%
  select(-iucn_sid) %>%
  mutate(deep = deep * 200) %>%
  distinct()

rast_coral_comp <- subs(x = loiczid_raster, y = coral_comp %>% select(loiczid, deep), 
                     by = 'loiczid', which = 'deep', 
                     subsWithNA = TRUE) %>%
  crop(extent(c(-180, 180, -63, 85)))

map_coral_comp <- 
  tm_shape(rast_coral_comp) +
    tm_raster(breaks = c(0, 200, Inf),
              palette = c('lightblue', 'royalblue4'),
              colorNA = NULL,
              title = 'depth (meters)',
              alpha = 1) +
  tm_shape(World) +
    tm_polygons(border.col = 'grey45', col = 'grey40', lwd = 0.25) + 
  tm_layout(basemaps = "Esri.WorldTopoMap", 
            legend.text.size = .35,
            legend.title.size = .45,
            # title.position = 'TOP', 
            legend.outside = FALSE, 
            legend.position = c('left', 'bottom'),
            legend.bg.color = 'white',
            legend.bg.alpha = .5,
            attr.outside = TRUE)

print(map_coral_comp)

save_tmap(map_coral_comp,
          filename = file.path(dir_fig, 'fig3a_coral_map.png'), 
          width = 9, height = 4.2, units = 'cm', dpi = 600)

### 2016-08-27 clip_fao/fao_clip_pac.Rmd
pushing aquamaps border issues and manuscript updates

hist(spp_quad_list$occurcells)
nrow(spp_quad_list %>% filter(occurcells > 100))
### 110 out of 1078 (~10%) have more than 100 occurrence cells

### load full AM list and filter to those in q3 and q4
am_spp_cells <- read_csv(file.path(dir_anx, 'spp_ico',
                                   'v2016/int/am_cells_spp_prob0.csv')) %>%
  filter(am_sid %in% spp_quad_list$am_sid) %>% 
  distinct()

### join with hcaf to get longitudes (and latitudes) for each cell
hcaf <- read_csv(file.path(dir_anx, '_raw_data/aquamaps/d2015',
                           'csv/hcaf_truncated.csv')) %>%
  select(-csquarecode)

am_spp_latlong <- am_spp_cells %>%
  left_join(hcaf, by = 'loiczid')

am_spp_latlong <- am_spp_latlong %>%
  mutate(clongadj  = ifelse(centerlong < 0, centerlong + 360, centerlong),
         wlimitadj = ifelse(wlimit < 0, wlimit + 360, wlimit),
         elimitadj = ifelse(elimit < 0, elimit + 360, elimit))

### filter out species from Atlantic... rough pass:
### - filter out everything west of 30 E and east of 70 W

am_spp_latlong <- am_spp_latlong %>%
  filter(clongadj >= 60 & clongadj <= 240)

@@

``` {r northmost_band}
### 5N north     - look for spike at 175 W (185E)
am_spp_north <- am_spp_latlong %>%
  filter(centerlat >= 5)

am_spp_north_lims <- am_spp_north %>%
  group_by(am_sid) %>%
  summarize(w_lim = min(wlimitadj),
            e_lim = max(elimitadj),
            n_w_lim = sum(wlimitadj == w_lim),
            n_e_lim = sum(elimitadj == e_lim),
            n_tot = n()) %>%
  gather(lim, long, w_lim:e_lim) %>%
  gather(lim_n, ncells, n_w_lim:n_e_lim) %>%
  filter(str_detect(lim_n, lim)) %>%
  select(-lim_n) %>%
  mutate(long = ifelse(long <= 70 | long >= 200, NA, long)) %>%
  filter(!is.na(long))
    ### delete artificially imposed limits

@@

am_spp_central_lims <- am_spp_central %>%
  group_by(am_sid) %>%
  summarize(w_lim = min(wlimitadj),
            e_lim = max(elimitadj),
            n_w_lim = sum(wlimitadj == w_lim),
            n_e_lim = sum(elimitadj == e_lim),
            n_tot = n()) %>%
  gather(lim, long, w_lim:e_lim) %>%
  gather(lim_n, ncells, n_w_lim:n_e_lim) %>%
  filter(str_detect(lim_n, lim)) %>%
  select(-lim_n) %>%
  mutate(long = ifelse(long <= 70 | long >= 200, NA, long)) %>%
  filter(!is.na(long))
    ### delete artificially imposed limits

@@

@@

stack_plot_both <- 
  tm_shape(mapstack_w) +
    tm_raster(palette = 'Reds',
              colorNA = NULL,
              title = 'n_spp',
              alpha = 1) +
  tm_shape(mapstack_e) +
    tm_raster(palette = 'Blues',
              colorNA = NULL,
              title = 'n_spp') +
  tm_shape(World) +
    tm_polygons() + 
  tm_shape(fao_polys) +
    tm_borders() +
  tm_layout(basemaps = "Esri.WorldTopoMap", 
            # title.position = 'TOP', 
            legend.outside = TRUE, attr.outside = TRUE)

>>>>

nrow(spp_quad_list %>% filter(occurcells > 100))
### 110 out of 1078 (~10%) have more than 100 occurrence cells

### load full AM list and filter to those in q3 and q4
am_spp_cells <- read_csv(file.path(dir_anx, 'spp_ico',
                                   'v2016/int/am_cells_spp_prob0.csv')) # %>%
  # filter(am_sid %in% spp_quad_list$am_sid) %>% 
  # distinct()

message('loading hcaf and adjusting to se asia')
### join with hcaf to get longitudes (and latitudes) for each cell
hcaf <- read_csv(file.path(dir_anx, '_raw_data/aquamaps/d2015',
                           'csv/hcaf_truncated.csv')) %>%
  select(-csquarecode, -cellarea, -oceanarea) %>%
  mutate(clongadj  = ifelse(centerlong < 0, centerlong + 360, centerlong),
         wlimitadj = ifelse(wlimit < 0, wlimit + 360, wlimit),
         elimitadj = ifelse(elimit < 0, elimit + 360, elimit)) %>%
  filter(clongadj >= 60 & clongadj <= 240)

### filter out species from Atlantic... rough pass:
### - filter out everything west of 30 E and east of 70 W
message('binding hcaf to am_spp_latlong')

am_spp_latlong <- hcaf %>%
  left_join(am_spp_cells, by = 'loiczid')

@@

``` {r northmost_band, eval = FALSE}
### 5N north     - look for spike at 175 W (185E)
am_spp_north <- am_spp_latlong %>%
  filter(centerlat >= 5)

am_spp_north_lims <- am_spp_north %>%
  group_by(am_sid) %>%
  summarize(w_lim = min(wlimitadj),
            e_lim = max(elimitadj),
            n_w_lim = sum(wlimitadj == w_lim),
            n_e_lim = sum(elimitadj == e_lim),
            n_tot = n()) %>%
  gather(lim, long, w_lim:e_lim) %>%
  gather(lim_n, ncells, n_w_lim:n_e_lim) %>%
  filter(str_detect(lim_n, lim)) %>%
  select(-lim_n) %>%
  filter(!(long <= 70 | long >= 200))
    ### delete artificially imposed limits

@@

message('summarizing...')
am_spp_central_lims <- am_spp_central %>%
  group_by(am_sid) %>%
  summarize(w_lim = min(wlimitadj),
            e_lim = max(elimitadj),
            n_w_lim = sum(wlimitadj == w_lim),
            n_e_lim = sum(elimitadj == e_lim),
            n_tot = n())

message('gathering...')
am_spp_central_lims <- am_spp_central_lims %>%
  gather(lim, long, w_lim:e_lim) %>%
  gather(lim_n, ncells, n_w_lim:n_e_lim) %>%
  filter(str_detect(lim_n, lim)) %>%
  select(-lim_n) %>%
  filter(!(long <= 70 | long >= 200))
    ### delete artificially imposed limits

write_csv(am_spp_central_lims, file.path(dir_git, 'clip_fao/am_spp_central_lims.csv'))

@@

clipped_spp <- read_csv(file.path(dir_git, 'clip_fao', 'clipped_mid_pac.csv'))

message('grouping affected species and summing to loiczid')
### group affected spp by loiczid and sum
spp_cells_clipped <- am_spp_cells %>%
  filter(am_sid %in% spp_examples_e$am_sid) %>%
  group_by(loiczid) %>%
  summarize(n_spp = n())

### create raster, then shift raster

rast_spp  <-  subs(loiczid_raster, 
                   spp_cells_clipped[ , c('loiczid', 'n_spp')], 
                   by = 'loiczid', 
                   which = 'n_spp', 
                   subsWithNA = TRUE)

rast_spp_shift <- rast_spp %>%
  raster::shift(360)

rast_spp_e <- merge(rast_spp, rast_spp_shift)

values(rast_spp_e)[values(rast_spp_e) == 0] <- NA
values(rast_spp_e)[is.na(values(rast_spp_e))] <- 0
rast_spp_e <- crop(rast_spp_e, extent(c(85, 220, -60, 60)))

rast_spp_map_e <- 
  tm_shape(rast_spp_e) +
    tm_raster(palette = 'Reds',
              breaks = c(1, 15, 100, 200, 300, Inf),
              colorNA = 'lightblue1',
              textNA = 'Zero',
              title = 'Species count',
              alpha = 1) +
  tm_shape(World) +
    tm_polygons(border.col = 'grey50', col = 'grey55') + 
  tm_shape(fao_polys) +
    tm_borders(col = 'grey40') +
  tm_layout(basemaps = "Esri.WorldTopoMap", 
            # title.position = 'TOP', 
            legend.outside = TRUE, attr.outside = TRUE)

print(rast_spp_map_e)

clipped_spp <- read_csv(file.path(dir_git, 'clip_fao', 'clipped_mid_pac.csv'))

@@

stack_plot_both <- 
  tm_shape(mapstack_w) +
    tm_raster(palette = 'Reds',
              colorNA = NULL,
              title = 'n_spp',
              alpha = 1) +
  tm_shape(mapstack_e) +
    tm_raster(palette = 'Reds',
              colorNA = NULL,
              title = 'n_spp') +
  tm_shape(World) +
    tm_polygons() + 
  tm_shape(fao_polys) +
    tm_borders() +
  tm_layout(basemaps = "Esri.WorldTopoMap", 
            # title.position = 'TOP', 
            legend.outside = TRUE, attr.outside = TRUE)

### 2016-08-27 ms_1_intro.Rmd
pushing aquamaps border issues and manuscript updates

One major outcome of this body of science is the various compiled databases of species distribution maps. Here we focus on global-scale repositories that predict marine species ranges throughout the world's oceans – AquaMaps [@kaschner_aquamaps:_2015] and International Union for Conservation of Nature (IUCN) [@iucn_iucn_2015].  These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].

___sentence about SAUP - how can we rephrase the "two global-scale repositories" sentence to exclude it? "broad taxonomic coverage"?  something about not focusing just on economically important species? (is that a valid assumption about SAUP) - consider other datasets as well: NatureServe, BOTW, etc___

___or: "Several global-scale repositories predict ... In this paper we focus on two datasets that cover a broad taxonomic range: ..."___

Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each offers an understanding of the "truth" - one relying primarily on expert opinion and the other on model predictions.  Each method presents its own challenges and advantages, and the small overlap in mapped species means each dataset provides value in taxonomic and geographic coverage.  Importantly, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: geographic range data such as IUCN range maps frequently introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of the `r format(n_spp_all, big.mark = ',')` marine species mapped by one or both datasets.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to marine biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique marine species).

As understanding of a given species improves, we expect that range maps, regardless of method, would become more accurate in predicting species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both distribution and overall area; from the results we determine methodological issues that can introduce commission errors, and suggest methods to improve confidence in species range predictions.

@@

Each of these datasets provides unique value in taxonomic and geographic coverage; while we have identified some potential issues in each, we cannot simply recommmend one data set over the other.  Instead, we recommend simple methods to reduce incidence of commission errors in each dataset, improving confidence in species range predictions to better inform conservation management and policy decisions.

>>>>

One major outcome of this body of science is the various compiled databases of species distribution maps. The two most comprehensive, widely-used global-scale repositories that predict marine species ranges throughout the world's oceans are AquaMaps [@kaschner_aquamaps:_2015] and range data from the International Union for Conservation of Nature (IUCN) [@iucn_iucn_2015].  Neither dataset claims to represent the "truth" of a species' spatial distribution, but rather each offers an understanding of the "truth", with IUCN ranges relying primarily on expert opinion and AquaMaps on model predictions.  Uncertainties inherent in method and intent inevitably drive differences in the range predictions made by each dataset: geographic range data such as IUCN range maps frequently introduce commission errors (inaccurate indication of presence), while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors (inaccurate indications of absence) [@rondinini_tradeoffs_2006].  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006; @jetz_ecological_2008].

These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status [@halpern_index_2012; @halpern_patterns_2015; @selig_assessing_2013-1], evaluating global biodiversity patterns [@coll_biodiversity_2010; @martin_manual_2014; @pimm_biodiversity_2014; @kaschner_current_2011], predicting species range shifts [@garciamolinos_climate_2015], and setting conservation priorities [@klein_shortfalls_2015].  In most of these cases, the implications of choosing one versus the other of these datasets is not evaluated or discussed, yet strong conservation and management conclusions are drawn from the results of the studies. For example, biases in taxonomic or spatial coverage within a dataset could shift management and conservation actions away from places or species that are most in need.

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of the `r format(n_spp_all, big.mark = ',')` marine species mapped by one or both datasets.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to marine biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique marine species).  Yet the small overlap in mapped species between the two datasets means each provides unique value in taxonomic and geographic coverage.

Although relatively small in number, these overlap species present a unique opportunity to evaluate the two datasets overall.  As understanding of a given species improves, we expect that range maps, regardless of method, would become more accurate in predicting species presence and absence.  Close alignment between two independent range maps can therefore indicate higher confidence in species presence or absence than a single map.  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align in both distribution and overall area; from the results we determine methodological issues that can introduce commission errors, and suggest methods to improve confidence in species range predictions.  In particular, we show how solving a simple problem in coral ranges in the IUCN range data dramatically ___improves agreement with Aquamaps data.___ - _does this make it sound like agreement is the goal? make it clear that agreement implies better confidence_

@@

While we have identified some potential issues in each dataset, we cannot simply recommend one data set over the other.  Instead, we recommend simple methods to reduce incidence of commission errors in each dataset, improving confidence in species range predictions to better inform conservation management and policy decisions.

### 2016-08-27 ms_3_res_disc_concl.Rmd
pushing aquamaps border issues and manuscript updates

@@

@@

@@

[results/discussion] ___area ratio suggests similar commission/omission errors, but dist suggests otherwise___ Species found in the lower right quadrant (n = `r n_q3`) seem to represent cases of "two wrongs make a right."  For these species, IUCN and AquaMaps both predict ranges extending far beyond the overlapping region, but the methodological differences result in very different extrapolations.   ___OMIT:___ _For these species, IUCN ranges frequently overextend into unsuitable depths, as in the case of many upper left quadrant species, while at the same time AquaMaps ranges often aggressively extrapolate presence into locations where IUCN predicts absence, likely introducing additional commission errors._  Consequently, area ratios are close to 100%, though the poor distribution alignment indicates that one or both datasets are introducing significant errors. As above, further analysis is required to disentangle the causes of error.

@@

## Case study: Coral depth analysis

@@

[results] Fig. 3A shows the 463 coral species mapped in the IUCN dataset, with their ranges broken into proportional area deeper and shallower than 200 m.  None of these species is indicated to occur deeper than 200 m, and 94% are confined to waters shallower than 50 m; seven of the mapped species had no reported depth information.  Clipping coral ranges to shallower than 200 m eliminated an average of 47.6% of the total predicted area while still allowing for a conservative estimate of suitable habitat.

>>>>

- figure 1 and caption

@@

## Distribution and range area alignment

@@

- quad plot and bar chart and captions

@@

[results/discussion] ___area ratio suggests similar commission/omission errors, but dist suggests otherwise___ Species found in the lower right quadrant (n = `r n_q3`) seem to represent cases of "two wrongs make a right."  For these species, IUCN and AquaMaps both predict ranges extending far beyond the overlapping region, but the methodological differences result in very different extrapolations.   Consequently, area ratios are close to 100%, though the poor distribution alignment indicates that one or both datasets are introducing significant errors. As above, further analysis is required to disentangle the causes of error.

@@

## Coral depth exploration

@@

[results] Fig. 3A shows the 463 coral species mapped in the IUCN dataset, with their ranges broken into proportional area deeper and shallower than 200 m.  ___show a map of all coral species aggregated, with deeper than 200 m cells shown___  None of these species is indicated to occur deeper than 200 m, and 94% are confined to waters shallower than 50 m; seven of the mapped species had no reported depth information.  Clipping coral ranges to shallower than 200 m eliminated an average of 47.6% of the total predicted area while still allowing for a generous estimate of suitable habitat.
